Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/overview.tex
...

\section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
\label{sec:example}

@@

\input{figures/quickselect}

Suppose that we wish to issue order statistics queries on a collection of items
represented as a list \texttt{l}. We can use quickselect \cite{Hoare:1961}, an
expected linear time algorithm which takes a list \texttt{l} and a rank
\texttt{k} and returns the element with rank \texttt{k} in \texttt{l}.

Quickselect, which we define in \ref{fig:qs-unstaged}, first partitions the list
by using the first element as a pivot%
\footnote{We assume that the list is prepermuted to guarantee the expected
linear time behavior.}
and then recurs on one of the two resulting sides to find the desired element.
The side chosen is determined by the relationship of \texttt{k} to the size of
the first half \texttt{n}, which is returned by the \texttt{part}ition function
along with the two sides themselves.

Now, consider an application where we perform many order statistics queries on
the same collection \texttt{l}, but with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$.
Certainly, it is possible to implement this with $m$ calls to
\texttt{quickselect}:
%
\begin{lstlisting}
quickselect l @$\mathtt{k_1}$@
quickselect l @$\mathtt{k_2}$@
 @$\vdots$@ 
quickselect l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Can we do better?

@@

An astute programmer might notice that much of the code in \texttt{quickselect}
does \emph{not} depend on the rank \texttt{k}; for example, the list is
partitioned before \texttt{k} is ever used. Moreover, while the recursive calls
to \texttt{qSelect} are guarded by a comparison to \texttt{k}, \texttt{k} does
not determine the arguments to those recursive calls---it only determines which
call is made. Thus, if we are willing to change the evaluation order of the
language (and evaluate under the \texttt{case} statement), it should be possible
to force \emph{all} the computations involving \texttt{l} to be performed before
\emph{any} of the computations involving \texttt{k}.

@@

Suppose now that $m$ is large (close to $n$ for example).  In this
case, our program is essentially a quadratic-time algorithm, which can
be very slow even for modest values of $m$ and $n$. In this case, to
improve the run-time it appears that we have no choice but rewrite our
program to use a different algorithm.  One way to do this would be to
pre-sort the input list $l$ and then lookup the element with the
desired rank (as specified by the second argument).  While this
approach might sound sensible, attaining the desired performance
requires answering one more important question: what should the data
structure for storing the sorted input be?  Unfortunately using a list
does not work, because each look up in a list requires linear time on
average. One option is to use arrays but in this case, we would need
to implement a binary search to perform the lookup.  Another option is
to use a balanced binary search tree data structure, which supports
lookups in logarithmic time.  Using either one of these techniques, we
can reduce the run-time to $O(n\log{n} + m\log{n})$, which is
significantly better that $O(n^2)$.

In summary, to achieve the desired improvement, we wrote a two-pass
program, where the first pass sorts the input and the second pass
performs a binary search either on an array or a balanced tree. As is
probably apparent to the reader, the resulting program is both
algorithmically and structurally more complex: it requires conversions
between different data structures (list to array or binary search
tree) and requires implementing additional algorithms such as sorting
and binary search, also possibly binary search trees.  As suggested by
this relatively small example and experienced by the graphics
community, the design, implementation, and maintenance of such
multi-pass programs can be difficult due to the increased complexity.

@@

When applied to the quick-select example as described, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation described above.  Specifically, in the first
pass, the program takes the input list and constructs a
probabilistically balanced binary search tree, which is isomorphic to
a treap data structure~\cite{treaps}.  In the second pass, the program
performs a binary search tree lookup, by walking the tree to find the
element with the desired rank.  The resulting program takes the
$O(n\log{n} + m\log{n})$.

We note that the splitting algorithm operates by composing local
transformations on the subterms of the input program.  In particular,
the algorithm has no special knowledge of the quick-select algorithm,
binary search trees, or how to perform lookups on binary search trees,
but it is able to derive all of these from the input program.  But
how?

The technical detail of the algorithm are relatively involved but in
the hope that it aids understanding we present a brief informal
overview of how the algorithm operates.  This exposition is
necessarily imprecise but the algorithm is made precise in
\ref{sec:splitting} and implemented
(\ref{sec:implementation}{sec:examples}).

At a high level, the splitting algorithm operates on our example by
using the code for quickselect and the knowledge that the list (the
first argument) is a first-stage argumint and the rank (second
argument) is a second stage argumunt, the algorithm then derives two
functions: \texttt{qsStage1} and \texttt{qsStage2} as follows.

\paragraph{Pass-1 code.}
The splitting algorithm ``imitates'' an execution of quick-select with
just the first-stage input by evaluating as much code as possible with
the available information.  Specifically the algorithm generates a
function \texttt{qsStage1} to perform all the recursive calls of
\texttt{qs} and to evaluate all instances of the partition function,
which depend only on the input list.  The function \texttt{qsStage1}
produces a trace of the execution of \texttt{qs} on its first input by
generating a tree that collects the results from all recursive calls
along with a tag that indicates the control branches taken.  The trace
ends up being represented as a tree because \texttt{qs} has a binary
control structure casing on the list. In fact, function
\texttt{qsStage1} constructs a binary search tree, keyed by the
``pivot'' used by the partition, because the ``pivot'' is needed in
the second stage to construct the result.  We call the result data
structure of \texttt{qsStace1} the {\em boundary}, as it passes
information from the first pass to the second.

\paragraph{Pass-2 code}
As the splitting algorithm ``scans'' \texttt{qs} for computations that
can be performed in the first stage and collects them into
\texttt{qsStage1}, it also collects computations that must be left to
the second stage in a separate function \texttt{qsStage2}.  This
function, which is executed in the second pass, takes as argument the
boundary and the second-stage argument (the rank) and performs a
lookup in the boundary data structure.  But how does \texttt{qsStage2}
knows to traverse the tree? It does not. Since the splitting algorithm
recorded the control flow of the first stage execution in the boundary
data structure, \texttt{qsStage2} simply follows this control
structure and performs at each point the parts of the computation from
\texttt{qs} that can now be performed in the presence of the
second-stage argument (the rank).

>>>>

\section{Overview}
\label{sec:examplea}

We present an overview of our approach by considering an example.  To
highlight the ideas, we abstract over some technical details, which
are fully formalized in the rest of the paper.  The example itself is
also implemented in our implementation.

\paragraph{Computing order statistics.}

@@

\input{figures/quickselect-aligned}

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which give a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected linear time.
%
\footnote{We assume that the list is prepermuted to guarantee the
  expected linear time behavior.} 
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example but with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate, because for example, because we wish to be
able to map \texttt{qs} to over another list and do so
efficiently.

One way to regain efficiency is to change the algorithm that we use to
perform the computation in multiple passes.  One option would be to
pre-sort the input list \texttt{l} into a sorted list \texttt{s} in
the first pass, and then perform lookups in \texttt{s} in the second
pass.  Unfortunately, since \texttt{s} is a list, a lookup would
require linear time (also on average), leading to no improvement in
efficiency.  We can resolve this issue by using a data structure that
can provide faster access.  For example, after sorting \texttt{s} we
can copy in into an array and then perform binary searches to find the
element with the desired rank (copying the input to an array and
sorting would not work well in a functional language).  As another
option, we can represent construct a binary search tree from the input
and perform size-based searches on the binary search tree to find the
elements with the desired rank.  Such an approach would fit nicely
into the functional paradigm of computing but requires additional
algorithmic ingenuity: to ensure efficiency we would need to store at
each note the size of its subtree (these are sometimes called
``augmented trees'').  In summary, we are able to improve efficiency
by replacing our simple solution with a multi-pass algorithm that
first preprocesses the input to generate a lookup data structure and
then in the second pass performs fast lookups.  As outlined above, in
addition, this transformation is highly non-trivial as it involves
reasoning about intricate algorithmic concerns, and implementing more
complex algorithms.

Such transformations, called pass-separation by J{\o}rring and
Scherlis are commonly employed.  For example, as briefly mentioned in
the introduction, modern graphics software is written exactly in this
way.  Unfortunately, as the example illustrates, they lead to complex
software, leading to the question of whether it would be possible to
perform such transformations automatically.  We answer this question
affirmatively.  Our approach is to start with a staged language and
then use a splitting algorithm to generate multi-pass programs from
staged programs.

@@

The idea behind staged programming is to use staging annotations or
types so indicate the stage of each expression.  To apply this
technique to our example, suppose that we have a language with two
stages.  We can now write a staged version of \texttt{qs}, called
\texttt{qss} in \ref{fig:qs-staged}, making the input list a stage-1
and the rank a stage-2, more precisely the input list has type
$\curr\mathrm{list}$ (an integer list ``now''), the rank has type
$\fut\rmint$ (an integer in the ``future''), and return has type
$\fut\rmint$.  We can then annotate the rest of the body of
\texttt{qs}, by using the staging annotations \texttt{prev} and
\texttt{next}, which respectively include a stage-1 expressions in
stage-2, and a stage-2 expression in stage-1.  To help readability, we
write the first-stage computations in red, and the second-stage
computations in blue.

the
function \texttt{qs} does \emph{not} depend on the rank \texttt{k};
for example, the list is partitioned before \texttt{k} is ever
used. Moreover, while the recursive calls to \texttt{qSelect} are
guarded by a comparison to \texttt{k}, \texttt{k} does not determine
the arguments to those recursive calls---it only determines which call
is made. Thus, if we are willing to change the evaluation order of the
language (and evaluate under the \texttt{case} statement), it should
be possible to force \emph{all} the computations involving \texttt{l}
to be performed before \emph{any} of the computations involving
\texttt{k}.

@@

Suppose now that $m$ is large (close to $n$ for example).  In this
case, our program is essentially a quadratic-time algorithm, which can
be very slow even for modest values of $m$ and $n$. In this case, to
improve the run-time it appears that we have no choice but rewrite our
program to use a different algorithm.  One way to do this would be to
pre-sort the input list $l$ and then lookup the element with the
desired rank (as specified by the second argument).  While this
approach might sound sensible, attaining the desired performance
requires answering one more important question: what should the data
structure for storing the sorted input be?  Unfortunately using a list
does not work, because each look up in a list requires linear time on
average. One option is to use arrays but in this case, we would need
to implement a binary search to perform the lookup.  Another option is
to implement a balanced binary search tree data structure for storing
the elements in the list and then use balanced binary-search-tree
lookups, which are logarithmic time.  Using either one of these
techniques, we can reduce the run-time to $\Theta(n\log{n} + m\log{n})$,
which is significantly better that $\Theta(n^2)$.

In summary, to achieve the desired improvement, we wrote a two-pass
program, where the first pass sorts the input and the second pass
performs a binary search either on an array or a balanced tree. As is
probably apparent to the reader, the resulting program is both
algorithmically and structurally more complex: it requires conversions
between different data structures (list to array or binary search
tree) and requires implementing additional algorithms such as sorting
and binary search or possibly binary search trees.

%% As suggested by this relatively small example and experienced for
%% example by the graphics community, the design, implementation, and
%% maintenance of such multi-pass programs can be difficult due to the
%% increased complexity.

@@

When applied to the quick-select example as described, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation outlined above.  Specifically, in the first pass,
the program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{treaps}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.  The resulting program, shown
in \ref{fig:????}  takes the $O(n\log{n} + m\log{n})$.

To create the multi-pass algorithm, the splitting algorithm operates
by composing local transformations on the subterms of the input
program.  In particular, the algorithm has no special knowledge of the
quick-select algorithm, binary search trees, or how to perform lookups
on binary search trees, but it is able to derive all of these from the
input program.  But how?
%
The details of the algorithm are relatively involved but in the hope
that it aids understanding we present a brief informal overview of how
the algorithm operates.  
%
%% This exposition is necessarily imprecise but the algorithm is made
%% precise in \ref{sec:splitting} and implemented
%% (\ref{sec:implementation}{sec:examples}).
%

At a high level, the splitting algorithm operates on our example by
using the program code and the knowledge that the list (the first
argument) is first-stage and the rank ( the second argument) is second
stage, the algorithm then derives two functions: \texttt{qsStage1} and
\texttt{qsStage2} as follows.

\paragraph{Pass-1 code.}
The splitting algorithm scans the program and imitates an execution of
\texttt{qs} with just the first-stage input by evaluating as much code
as possible with the available first-stage values.  Specifically the
algorithm generates a function \texttt{qsStage1} to perform all the
recursive calls of \texttt{qs} and to evaluate all instances of the
partition function, which depend only on the input list.  The function
\texttt{qsStage1} produces a trace of the execution of \texttt{qs} on
its first input by generating a tree that collects the results from
all recursive calls along with a tag that indicates the control
branches taken.  The trace is represented as a tree because
\texttt{qs} has a binary control structure (casing on the list), which
leads to a binary tree data structure.  To ensure that the trace
contains the necessary results to complete the execution in the second
pass, where second-stage values may be used, the function
\texttt{qsStage1} includes in the nodes of the tree information such
as the key used for splitting the list at that recursive call.  The
resulting tree is thus a binary search tree, keyed by the ``pivot''
used by the partition.  In general we refer to the result data
structure of \texttt{qsStace1} as the {\em boundary}, as it passes
information from the first pass to the second.

\paragraph{Pass-2 code}
As the splitting algorithm scans \texttt{qs} for computations that can
be performed in the first stage and collects them into
\texttt{qsStage1}, it also collects computations that must be left to
the second stage in a separate function \texttt{qsStage2}.  This
function, which is executed in the second pass, takes as argument the
boundary and the second-stage argument (the rank) and performs a
lookup in the boundary data structure.  But how does \texttt{qsStage2}
knows to traverse the tree? It does not. Since the splitting algorithm
recorded the control flow of the first stage execution in the boundary
data structure, \texttt{qsStage2} simply follows this control
structure and performs at each point the parts of the computation from
\texttt{qs} that can now be performed in the presence of the
second-stage argument (the rank).  In the context of our example, this
performs a lookup on the boundary data structure by using the supplied
rank.

\paragraph{Summary.}
In summary, the splitting algorithm scans the program code for
first-stage computations (which depend only on first-stage values) and
separates them into function of the first pass. This function performs
the stage-1 computations and places the results into a boundary data
structure that both records the control flow and the results from the
first stage computation at each control-flow point.  As the splitting
function generates the code for the first pass, it also generates the
code for the second pass as a function that walks both the boundary
data (which essentially imitates execution) and at each control point
performs the part of the computation that depends on second-stage
values. We evaluate the resulting program by first evaluating the
function for the first pass and passing its return value (the
boundary) to the second pass, which then returns the desired result.
